The Perils of Exploration under Competition: \\ A Computational Modeling Approach

### 2021-03-02 journal paper/content/conclusion.tex
remove old color-edits

We study the tension between exploration and competition. We consider a stylized duopoly model in which two firms face an identical multi-armed bandit problem and compete for a stream of users. A firm makes progress on its learning problem if and only if it attracts users.
%\asdelete{We investigate two variants: an analytical variant where users do not observe any performance signals and choose between firms according to their Bayesian-expected rewards, and numerical simulations where a reputation score is observed for each firm, based on the average reward of its recent users.  In both variants,}
We find that firms are incentivized to adopt a ``greedy algorithm" which does no purposeful exploration and leads to welfare losses for users. We then consider two relaxations of competition: we soften users' decision rule and give one of the firms a first-mover advantage. Both relaxations induce firms to adopt ``better" bandit algorithms, which benefits user welfare.

>>>>

We study the tension between exploration and competition. We consider a stylized duopoly model in which two firms face the same multi-armed bandit problem and compete for a stream of users. A firm makes progress on its learning problem only if it attracts users.
%\asdelete{We investigate two variants: an analytical variant where users do not observe any performance signals and choose between firms according to their Bayesian-expected rewards, and numerical simulations where a reputation score is observed for each firm, based on the average reward of its recent users.  In both variants,}
We find that firms are incentivized to adopt a ``greedy algorithm" which does no purposeful exploration and leads to welfare losses for users. We then consider two relaxations of competition: we soften users' decision rule and give one firm a first-mover advantage. Both relaxations induce firms to adopt ``better" bandit algorithms, which benefits user welfare.

### 2021-03-02 journal paper/content/related_work.tex
remove old color-edits

\cite{schmalensee1982product, bagwell1990informational} investigate how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers; we find a similar effect with ``reputation advantage". \cite{de2020data} note the role of data as a barrier to entry in online markets; we find a similar effect with ``data advantage". \citet{kerin1992first} overview other channels through which first-mover advantage can \asedit{affect competition}.% lead to a competitive advantage.

@@

\asedit{While we use first-mover advantage and agents' decision rule,} classic ``market competitiveness" measures, such as the Lerner Index or the Herfindahl-Hirschman Index
\citep{tirole1988theory}, are not applicable to our setting, as they rely on ex-post observable market attributes such as prices or market shares (which are, resp., absent and endogenous for us).

>>>>

\cite{schmalensee1982product, bagwell1990informational} investigate how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers; we find a similar effect with ``reputation advantage". \cite{de2020data} note the role of data as a barrier to entry in online markets; we find a similar effect with ``data advantage". \citet{kerin1992first} overview other channels through which first-mover advantage can affect competition.% lead to a competitive advantage.

@@

While we use first-mover advantage and agents' decision rule, classic ``market competitiveness" measures, such as the Lerner Index or the Herfindahl-Hirschman Index
\citep{tirole1988theory}, are not applicable to our setting, as they rely on ex-post observable market attributes such as prices or market shares (which are, resp., absent and endogenous for us).

### 2021-03-02 journal paper/content/sec-intro.tex
remove old color-edits

While traditional models of innovation study lab-based R\&D, we consider data-driven innovation, which crucially depends on data generated by the firm's customers.
We focus on innovation in \emph{exploration technology} which systematically improves the firm's products, whereas prior work would define innovation as improvement in the products themselves. We recover the \asdelete{traditional} inverted-U relationship purely through the reputational consequences of exploration.
\asdelete{rather than monetary costs/benefits thereof.}
By contrast, the inverted-U relationships from prior work rely on the monetary aspects: \asedit{investments into  R\&D and profits from innovation.}

@@

%\begin{itemize}
\textbf{(i)}
%We compare \HardMax and \HardMaxRandom decision rules.
We find that the greedy algorithm often wins under the \HardMax decision rule, with a strong evidence of the ``death spiral" effect mentioned earlier. As predicted by the theory, better algorithms prevail under \HardMaxRandom
\asedit{with enough ``random" users.}
%if the expected number of ``random" users is sufficiently large.
 %\asdelete{However, this effect is negligible for smaller parameter values.}

@@

\textbf{(iii)}
However, the greedy algorithm is sometimes \emph{not} the best strategy under high levels of competition.\footnote{In our theoretical results on \HardMax, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations  the prior is not available).}
We revisit algorithms' performance in a standalone bandit problem, \ie without competition. We find that the most natural performance measure does not explain this phenomenon, and suggest a new, more nuanced \asedit{one} that does.
%performance measure that does.

@@

Interestingly,
\asdelete{we find that}
data feedback loops can be bad for innovation (as in ``death spiral"), or good for innovation (as in ``amplification effect"), depending on the level of competition.

>>>>

While traditional models of innovation study lab-based R\&D, we consider data-driven innovation, which crucially depends on data generated by the firm's customers.
We focus on innovation in \emph{exploration technology} which systematically improves the firm's products, whereas prior work would define innovation as improvement in the products themselves. We recover the traditional inverted-U relationship purely through the reputational consequences of exploration.
\asdelete{rather than monetary costs/benefits thereof.}
By contrast, the inverted-U relationships from prior work rely on the monetary aspects: \asedit{investments into  R\&D and profits from innovation.}

@@

%\begin{itemize}
\textbf{(i)}
%We compare \HardMax and \HardMaxRandom decision rules.
We find that the greedy algorithm often wins under the \HardMax decision rule, with a strong evidence of the ``death spiral" effect mentioned earlier. As predicted by the theory, better algorithms prevail under \HardMaxRandom with enough ``random" users.
%if the expected number of ``random" users is sufficiently large.
 %\asdelete{However, this effect is negligible for smaller parameter values.}

@@

\textbf{(iii)}
However, the greedy algorithm is sometimes \emph{not} the best strategy under high levels of competition.\footnote{In our theoretical results on \HardMax, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations  the prior is not available).}
We revisit algorithms' performance in a standalone bandit problem, \ie without competition. We find that the most natural performance measure does not explain this phenomenon, and suggest a new, more nuanced one that does.
%performance measure that does.

@@

Interestingly, data feedback loops can be bad for innovation (as in ``death spiral"), or good for innovation (as in ``amplification effect"), depending on the level of competition.

### 2021-03-02 journal paper/content/sec-model.tex
remove old color-edits

\begin{OneLiners}
\item In the \emph{\TheoryModel}, the mean reward vector $\mu = (\mu_a:\; a\in A)$ is drawn from a common Bayesian prior $\priorMu$. Each agent knows its global round $t$, but not the performance signals such as the current market shares. Her reward estimates are defined as posterior mean rewards:
        $\est_i(t) = \PMR_i(t) := \E\sbr{ r_t\mid i_t = i}$
    for each principal $i$, \asedit{where the agent knows $t$, the principals' algorithms, $\priorMu$, and $\respF$.}
%\footnote{Agents do the Bayesian update knowing $t$, the principals' algorithms, the prior $\priorMu$, and $\respF$.}

@@

\xhdr{Competition game.}
Some of our results explicitly study the game between the two principals, termed the \emph{competition game}. We model it as a simultaneous-move game: before the first agent arrives, each principal commits to an MAB algorithm. (Since MAB algorithms can be randomized, we make no distinction between pure and mixed strategies.) Principals are risk-neutral; their utility is defined as their market share, \ie the number of agents \asedit{they attract}. Thus, they aim to select the algorithm that maximizes their expected market share.

>>>>

\begin{OneLiners}
\item In the \emph{\TheoryModel}, the mean reward vector $\mu = (\mu_a:\; a\in A)$ is drawn from a common Bayesian prior $\priorMu$. Each agent knows its global round $t$, but not the performance signals such as the current market shares. Her reward estimates are defined as posterior mean rewards:
        $\est_i(t) = \PMR_i(t) := \E\sbr{ r_t\mid i_t = i}$
    for each principal $i$, where the agent knows $t$, the principals' algorithms, $\priorMu$, and $\respF$.
%\footnote{Agents do the Bayesian update knowing $t$, the principals' algorithms, the prior $\priorMu$, and $\respF$.}

@@

\xhdr{Competition game.}
Some of our results explicitly study the game between the two principals, termed the \emph{competition game}. We model it as a simultaneous-move game: before the first agent arrives, each principal commits to an MAB algorithm. (Since MAB algorithms can be randomized, we make no distinction between pure and mixed strategies.) Principals are risk-neutral; their utility is defined as their market share, \ie the number of agents they attract. Thus, they aim to select the algorithm that maximizes their expected market share.

### 2021-03-02 journal paper/content/sec-theory.tex
remove old color-edits

We consider agents with \HardMax response function, and show that principals are not incentivized to \emph{explore}, \ie to deviate from \DynGreedy. The core technical result is that if one principal adopts \DynGreedy, then the other principal loses all agents as soon as he deviates therefrom. \asedit{We make this formal below.}

@@

Thus, the greedy modification is beneficial in competition \emph{and} keeps us inside the set of \bmonotone MAB algorithms. Consequently, if the principals are restricted \asedit{to the latter set,}
%to choosing \bmonotone MAB algorithms,
then non-greedy algorithms cannot be chosen in equilibrium.

@@

We consider the marginal utility of switching to a better algorithm. Suppose initially both principals use some algorithm \alg, and principal 1 ponders switching to another algorithm \alg' which BIR-dominates \alg. \asedit{What is the marginal utility $\deltaU$ of this switch?}

@@

\xhdr{Beliefs.}
Instead of knowing the principals' algorithms $(\alg[1],\alg[2])$, the Bayesian prior $\priorMu$, and the response function $\respF$, agents could have beliefs on these objects that need not be correct. If agents have common ``point beliefs" %respectively, algorithms $(\alg[1]',\alg[2]')$ prior $\priorMu'$ and response function $\respF'$,
\asedit{on these objects}, then all our results carry over with respect to these beliefs.

>>>>

We consider agents with \HardMax response function, and show that principals are not incentivized to \emph{explore}, \ie to deviate from \DynGreedy. The core technical result is that if one principal adopts \DynGreedy, then the other principal loses all agents as soon as he deviates therefrom. We make this formal below.

@@

Thus, the greedy modification is beneficial in competition \emph{and} keeps us inside the set of \bmonotone MAB algorithms. Consequently, if the principals are restricted to the latter set,
%to choosing \bmonotone MAB algorithms,
then non-greedy algorithms cannot be chosen in equilibrium.

@@

We consider the marginal utility of switching to a better algorithm. Suppose initially both principals use some algorithm \alg, and principal 1 ponders switching to another algorithm \alg' which BIR-dominates \alg. What is the marginal utility $\deltaU$ of this switch?

@@

\xhdr{Beliefs.}
Instead of knowing the principals' algorithms $(\alg[1],\alg[2])$, the Bayesian prior $\priorMu$, and the response function $\respF$, agents could have beliefs on these objects that need not be correct. If agents have common ``point beliefs" %respectively, algorithms $(\alg[1]',\alg[2]')$ prior $\priorMu'$ and response function $\respF'$,
on these objects, then all our results carry over with respect to these beliefs.

### 2021-03-02 journal paper/content/sim_details.tex
remove old color-edits

\xhdr{Simulation details.}
For each MAB instance we draw $N = 1000$ \MRVs independently from the corresponding distribution. We use this same collection of \MRVs for all experiments with this MAB instance. For each \MRV we draw a table of realized rewards (\emph{realization table}), and use this same table for all experiments on this \MRV. This ensures that differences in algorithm performance are not due to noise in the realizations but due to differences in the algorithms in the different experimental settings.

>>>>

\xhdr{Simulation details.}
For each MAB instance we draw $N = 1000$ \MRVs independently from the corresponding distribution. We use this same collection of \MRVs for all experiments with this MAB instance. For each \MRV we draw a table of realized rewards (\emph{realization table}), and use this same table for all experiments on this \MRV. This ensures that differences in algorithm performance are solely due to differences in the algorithms in the different experimental settings.

### 2021-03-02 journal paper/ec19paper/content/revisited.tex
more space saving edits
remove old color-edits

\begin{finding}
\textit{
Mean reputation trajectories do not suffice to explain the outcomes under competition.}
\end{finding}

@@

We see that the ``naive" algorithms $\DynamicGreedy$ and $\DynamicEpsGreedy$ have a bi-modal reputation distribution, whereas $\Thompson$ does not. The reason is that for this MAB instance, $\DynamicGreedy$ either finds the best arm and sticks to it, or gets stuck on the bad arms. In the former case \DynamicGreedy does slightly better than $\Thompson$, and in the latter case it does substantially worse. However, the mean reputation trajectory may fail to capture this complexity since it simply takes average over different \MRVs. This may be inadequate for explaining the outcome of the duopoly game, given that the latter is determined by a simple comparison between the firm's reputation scores.

>>>>

\begin{finding}
\textit{
Mean reputation trajectories do not explain the outcomes under competition.}
\end{finding}

@@

We see that the ``naive" algorithms $\DynamicGreedy$ and $\DynamicEpsGreedy$ have a bi-modal reputation distribution, whereas $\Thompson$ does not. The reason is that for this MAB instance, $\DynamicGreedy$ either finds the best arm and sticks to it, or gets stuck on the bad arms. In the former case \DynamicGreedy does slightly better than $\Thompson$, and in the latter case it does substantially worse. However, the mean reputation trajectory fails to capture this complexity since it takes average over different \MRVs. This is inadequate for explaining the outcome of the duopoly game, given that the latter is determined by a comparison between the firm's reputation scores.
